---
title: "Composite Floor Price Index (CFPI) for Palay"
subtitle: "Methodology and Implementation framework  for trigger Mechanism for Floor Prices"
author: "Data Analytics Center"
institute: "Philippine Rice Research Institute"
# format: revealjs
format:
  revealjs:
    css: www/custom-styles.css
    touch: true
    controls: true
    slide-number: true
    pdf-separate-fragments: true
    transition: fade
    chalkboard: 
      buttons: false
    logo: www/dac.png
    footer: "PhilRice Data Analytics Center"
execute: 
  echo: false
  warning: false
  message: false
  freeze: auto
server: shiny
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(DT)
library(htmltools)
library(plotly)
library(jsonlite)
library(httr)
library(DiagrammeR)
library(lmtest)
library(knitr)
library(purrr)
library(daclakeapi)
```

## 1. Introduction

-   **Executive Order No. 100 (2025)** mandates the establishment of a *floor price for palay* and the creation of *trigger mechanisms* for its timely implementation.
-   This policy recognizes that **volatile market prices**, especially during **peak harvest seasons**, expose farmers to unfair trading conditions, eroding incomes and discouraging continued rice production.
-   Rising production costs, climate-induced risks, and widening gaps between **farmgate and retail prices** further highlight the need for a **data-driven safeguard mechanism** that anticipates market downturns.

------------------------------------------------------------------------

## 2. Policy Context {.extra-tiny-text}

> “The government shall guarantee a fair and just return to farmers by setting a floor price that covers production costs and provides reasonable income, while ensuring affordability and stability of rice supply for consumers.”\
> — *Executive Order No. 100, 2025*

-   The **Department of Agriculture (DA)**, through the **Steering Committee**, is mandated to:
    -   Determine and adjust the floor price of palay;
    -   Monitor market signals and supply conditions;
    -   Recommend timely interventions to prevent farmgate price collapse.
-   The **DA** shall determine, set and adjust floor price of palay to be purchased by the government based on the following conditions:
    a.  Cost of Production
    b.  Prevailing market prices
    c.  Reasonable margins for farmers
    d.  Welfare of both farmers and consumers; and
    e.  Other relevant factors and conditions, such as:
        i.  Emergencies or calamities affecting the rice supply chain and palay production
        ii. Influx of import arrivals that could dampen farmgate price of palay
        iii. Sharp decline in global rice prices that may impact the estimated landed cost of rice; and
        iv. Government's capacity to purchase pa/ay at competitive prices and manage market intervention

------------------------------------------------------------------------

## 3. CFPI Design Principles

1.  Timeliness: Detect early signals of price decline (1-month lead).

2.  Comprehensiveness: Capture considerations mentioned in the EO.

3.Transparency: Computable and explainable index.

4.  Adaptability: Adjustable weights and thresholds for backtesting.

------------------------------------------------------------------------

## 4. CPFI Parameters {.tiny-text .scrollable}

::: extra-tiny-text
| **Category**                             | **Monthly Indicator**                                  | **Details / Formula**                                                                                                                                                   | **EO 100 Sec. 1 Considerations**                                    | **Relation to Farmer Risk**                               |      **Lag (Months)**       | **Source**             |
|:----------|:----------|:----------|:----------|:----------|:---------:|:----------|
| **A. FARM-LEVEL PRICES AND PROFIT**      | **1. Farmgate Price Deviation (**$FarmDev$)            | Deviation of farmgate price from monthly historical mean.<br>$$ FarmDev_t =\frac{FP_t - FP_{hist}}{FP_{hist}} $$                                                        | Prevailing market prices                                            | ↓ Lower deviation implies falling farmgate prices         |              1              | PSA                    |
|                                          | **2. Farmer Profit Deviation (**$ProfitDev$)           | Deviation of monthly farmer profit from its historical mean.<br>$$ Profit_t = FP_t - PCost_t $$ $$ ProfitDev_t = \frac{Profit_t - Profit_{hist}}{Profit_{hist}} $$      | Cost of production; Reasonable farmer margins                       | ↓ Declining profits increase farmer vulnerability         |              1              | PSA, PhilRice DAC      |
| **B. RETAIL-LEVEL PRICES AND PRICE GAP** | **3. Retail Price Deviation (**$RetDev$)               | Deviation of retail price from its historical monthly mean.<br>$$ RetDev_t =\frac{RP_t - RP_{hist}}{RP_{hist}} $$                                                       | Prevailing market prices; Welfare of farmers & consumers            | ↓ Signals weak consumer-side demand pressure              |              1              | PSA                    |
|                                          | **4. Retail–Farmgate Price Gap Deviation (**$RetGap$)  | Difference between retail and farmgate prices.<br>$$ Margin_t = RP_t - FP_t $$ $$RetGap_t = \frac{Margin_t - Margin_{hist}}{Margin_{hist}} $$                           | Reasonable margins for farmers; Welfare of both farmers & consumers | ↑ Rising gap may indicate farmer disadvantage             |              1              | PSA, PhilRice DAC      |
| **C. SUPPLY AND DAMAGES**                | **5. Area Planted / Expected Area Harvested (**$Area$) | Ratio of expected area to be harvested in month *t* to total semester area, adjusted for damaged area.<br> $$ Area_t = \frac{A^{harv}_t - A^{dam}_t}{A^{harv}_{sem}} $$ | Calamities, pest damage, and other supply shocks                    | ↑ Reduced harvest area signals production risk            | 0 (Current month preferred) | PRiSM, DA-FOS          |
| **D. COST FACTORS**                      | **6. Fertilizer Cost Deviation (**$FertDev$)           | Deviation of fertilizer prices (e.g., Urea, Complete) from historical monthly average.<br>$$ FertDev_t =\frac{FC_t - FC_{hist}}{FC_{hist}} $$                           | Reasonable farmer margins; Cost of production                       | ↑ Rising input costs compress farmer income               |              1              | FPA                    |
|                                          | **7. Fuel Price Deviation (**$FuelDev$)                | Deviation of monthly fuel price from 3-week moving average.<br>$$ FuelDev_t =\frac{FuelP_t - FuelP_{MA(3)}}{FuelP_{MA(3)}} $$                                           | Reasonable margin for farmers; Production logistics                 | ↑ Higher fuel costs increase production and hauling costs |              1              | DOE                    |
| **E. CLIMATE AND QUALITY FACTORS**       | **8. Monthly Rainfall Deviation (**$RainDev$)          | Deviation of observed monthly rainfall from historical mean.<br>$$ RainDev_t ==\frac{Rain_t - Rain_{hist}}{Rain_{hist}} $$                                              | Calamities or extreme weather affecting rice supply                 | ↑ Excess or deficient rainfall affects yield & quality    |              1              | PAGASA, CHIRPS, Others |
| **F. TRADE AND MARKET FACTORS**          | **9. Import Arrival Deviation (**$ImportDev$)          | Deviation of rice import arrivals from 3-month moving average.<br> $$ ImportDev_t =\frac{ImportQ_t - ImportQ_{MA(3)}}{ImportQ_{MA(3)}}$$                                | Influx of imports that could dampen farmgate price                  | ↑ More imports increase market supply pressure            |              1              | BOC, BPI               |
|                                          | **10. Global Rice Price Deviation (**$GlobalPriceDev$) | Deviation of average FOB of Viet 25% and.or Thai 25% rice from 3-month moving average.<br>$$ GlobalPriceDev_t = \frac{FOB_t -FOB_{MA(3)}}{FOB_{MA(3)}} $$               | Sharp decline in global rice prices may reduce local farmgate price | ↓ Lower global price increases downward pressure          |              1              | FAO, USDA, VFA         |
:::

------------------------------------------------------------------------

## Legend {.extra-tiny-text}

::: {.tiny-text .scrollable}
| **Symbol**         | **Definition / Description**                                                          | **Typical Unit** |
|:-------------|:--------------------------------------------|:------------:|
| $FP_t$             | Farmgate price in month *t*                                                           |       ₱/kg       |
| $FP_{hist}$        | Historical mean farmgate price for the same month (multi-year average)                |       ₱/kg       |
| $FarmDev_t$        | Farmgate price deviation from historical mean                                         |   \% or ratio    |
| $PCost_t$          | Production cost in month *t*                                                          |       ₱/kg       |
| $Profit_t$         | Farmer profit in month *t*, computed as $FP_t - PCost_t$                              |       ₱/kg       |
| $Profit_{hist}$    | Historical mean of monthly farmer profit                                              |       ₱/kg       |
| $ProfitDev_t$      | Deviation of current profit from historical mean                                      |   \% or ratio    |
| $RP_t$             | Retail price in month *t*                                                             |       ₱/kg       |
| $RP_{hist}$        | Historical mean retail price for the same month                                       |       ₱/kg       |
| $RetDev_t$         | Retail price deviation from historical mean                                           |   \% or ratio    |
| $Margin_t$         | Retail–farmgate price gap at month *t*, computed as $RP_t - FP_t$                     |       ₱/kg       |
| $Margin_{hist}$    | Historical mean of monthly retail–farmgate price gap                                  |       ₱/kg       |
| $RetGap_t$         | Deviation of current retail–farmgate price gap from historical mean                   |   \% or ratio    |
| $A^{harv}_t$       | Expected area harvested in month *t*                                                  |        ha        |
| $A^{dam}_t$        | Damaged area due to calamities/pests in month *t*                                     |        ha        |
| $A^{harv}_{sem}$   | Total expected harvested area for the semester                                        |        ha        |
| $Area_t$           | Ratio of effective harvested area in month *t* to total semester area                 |      ratio       |
| $FC_t$             | Fertilizer cost (e.g., Urea, Complete\$ in month *t*                                  |  ₱/kg or ₱/bag   |
| $FC_{hist}$        | Historical mean fertilizer cost for the same month                                    |  ₱/kg or ₱/bag   |
| $FertDev_t$        | Fertilizer price deviation from historical mean                                       |   \% or ratio    |
| $FuelP_t$          | Fuel price in month *t*                                                               |     ₱/liter      |
| $FuelP_{MA(3)}$    | 3-month moving average of fuel price                                                  |     ₱/liter      |
| $FuelDev_t$        | Fuel price deviation from 3-month moving average                                      |   \% or ratio    |
| $Rain_t$           | Observed monthly rainfall                                                             |        mm        |
| $Rain_{hist}$      | Historical mean rainfall for the same month                                           |        mm        |
| $RainDev_t$        | Deviation of observed rainfall from historical mean                                   |   \% or ratio    |
| $ImportQ_t$        | Volume of rice imports in month *t*                                                   |   metric tons    |
| $ImportQ_{MA(3)}$  | 3-month moving average of rice import volume                                          |   metric tons    |
| $ImportDev_t$      | Deviation of import volume from its 3-month moving average                            |   \% or ratio    |
| $FOB_t$            | Monthly average Free-on-Board (FOB) price of imported rice (e.g., Viet 25%, Thai 25%) |     USD/ton      |
| $FOB_{MA(3)}$      | 3-month moving average of global FOB price                                            |     USD/ton      |
| $GlobalPriceDev_t$ | Deviation of global rice price from its 3-month moving average                        |   \% or ratio    |
:::

------------------------------------------------------------------------

## 5. CFPI Computation {.tiny-text}


### Composite Floor Price Index (CFPI) Formula

$$
\text{CFPI}_t = \sum_{i=1}^{n} w_i \text{IND'}_{i,t}
$$

::: {.tiny-text .scrollable}

### Where:

| Symbol | Definition |
|:-------|:------------|
| $\text{CFPI}_t$| Composite Floor Price Index at time $t$, indicating market risk of farmgate price decline at time $t+1$ |
| $n$| Total number of indicators considered in the index |
| $i$| Indicator index (e.g., farmgate price deviation, profit deviation, etc.) |
| $w_i$| Assigned weight of the $i^{th}$indicator, reflecting its relative importance to farmer welfare or market risk |
| $\text{IND'}_{i,t}$| Value of indicator $i$ at time $t$, with directional correction, whenever applicable |

### Notes:
- A **higher CFPI value** suggests **increasing farm-level vulnerability** and may **trigger** setting or enforcing a floor price.  
- Indicator weights $w_i$can be determined using **expert elicitation**, **principal component analysis (PCA)**, or **equal weighting** as baseline.
- Directional correction  is applied to indicator that has an expected **inverse relationship** with farmer's risk and/or farmgate prices. For such indicators, the corrected value is obtained as $\text{IND'}_{i,t} = IND_{i,t} \times (-1)$.

:::

---

## 6. Condensed Formula {.tiny-text}

### Composite Floor Price Index (CFPI)

The CFPI is computed as the weighted sum of ten normalized and directionally corrected indicators representing key aspects of rice market dynamics, farmer profitability, and production risk.

$$
\begin{aligned}
\text{CFPI}_t =\;& 
w_1(FarmDev_t) +
w_2(ProfitDev_t) +
w_3(RetDev_t) +
w_4(RetGap_t) +
w_5(Area_t) + \\[6pt]
& w_6(FertDev_t) +
w_7(FuelDev_t) +
w_8(RainDev_t) +
w_9(ImportDev_t) +
w_{10}(GlobalPriceDev_t)
\end{aligned}
$$


### With direction correction

$$
\begin{aligned}
\text{CFPI}_t =\;& 
w_4(RetGap_t) +
w_5(Area_t) +
w_6(FertDev_t) +
w_7(FuelDev_t) +
w_8(RainDev_t) + \\[6pt]
&w_9(ImportDev_t) 
-w_1(FarmDev_t) -
w_2(ProfitDev_t) -
w_3(RetDev_t)  - 
w_{10}(GlobalPriceDev_t)
\end{aligned}
$$


---

## 7. Full Formula {.tiny-text}

### Composite Floor Price Index (CFPI)


$$
\begin{aligned}
\text{CFPI}_t &= 
w_1 \left(\frac{FP_t - FP_{hist}}{FP_{hist}}\right)
+ w_2 \left(\frac{Profit_t - Profit_{hist}}{Profit_{hist}}\right)
+ w_3 \left(\frac{RP_t - RP_{hist}}{RP_{hist}}\right) \\
&\quad + w_4 \left(\frac{Margin_t - Margin_{hist}}{Margin_{hist}}\right)
+ w_5 \left(\frac{A^{harv}_t - A^{dam}_t}{A^{harv}_{sem}}\right)
+ w_6 \left(\frac{FC_t - FC_{hist}}{FC_{hist}}\right) \\
&\quad + w_7 \left(\frac{FuelP_t - FuelP_{MA(3)}}{FuelP_{MA(3)}}\right)
+ w_8 \left(\frac{Rain_t - Rain_{hist}}{Rain_{hist}}\right)
+ w_9 \left(\frac{ImportQ_t - ImportQ_{MA(3)}}{ImportQ_{MA(3)}}\right) \\
&\quad + w_{10} \left(\frac{FOB_t - FOB_{MA(3)}}{FOB_{MA(3)}}\right)
\end{aligned}
$$


---


### With direction correction  {.tiny-text}

::: {.tiny-text}
$$
\begin{aligned}
\text{CFPI}_t &= 
w_4 \left(\frac{Margin_t - Margin_{hist}}{Margin_{hist}}\right)
+ w_5 \left(\frac{A^{harv}_t - A^{dam}_t}{A^{harv}_{sem}}\right)
+ w_6 \left(\frac{FC_t - FC_{hist}}{FC_{hist}}\right) \\
&\quad + w_7 \left(\frac{FuelP_t - FuelP_{MA(3)}}{FuelP_{MA(3)}}\right)
+ w_8 \left(\frac{Rain_t - Rain_{hist}}{Rain_{hist}}\right)
+ w_9 \left(\frac{ImportQ_t - ImportQ_{MA(3)}}{ImportQ_{MA(3)}}\right) \\
&\quad - w_1 \left(\frac{FP_t - FP_{hist}}{FP_{hist}}\right)
- w_2 \left(\frac{Profit_t - Profit_{hist}}{Profit_{hist}}\right)
- w_3 \left(\frac{RP_t - RP_{hist}}{RP_{hist}}\right) \\
&\quad - w_{10} \left(\frac{FOB_t - FOB_{MA(3)}}{FOB_{MA(3)}}\right)
\end{aligned}
$$
:::

---

## 8. CFPI Calculator tool

```{r}
shinyAppDir('cfpi_simulate')
```

